---
title: "Cyclistic RMD file"
author: "Saud"
date: "2024-11-26"
output: html_document
---

```{r}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
```

# Load required packages (ensure these are installed beforehand)
```{r}
library(tidyverse)
library(skimr)
library(janitor)
library(here)
library(scales)
library(conflicted)
```

# Resolve conflicts between packages
```{r}
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

# Download CSV files here [Trips Data](https://divvy-tripdata.s3.amazonaws.com/index.html)

# Load data
# Ensure the CSV files are downloaded into your RStudio working directory
```{r}
q1_2019 <- read_csv("Trips_2019_Q1.csv")
q1_2020 <- read_csv("Trips_2020_Q1.csv")
```

# Quick look at the datasets (avoid View to prevent GUI-related issues)
```{r}
# Display the first few rows of each dataset
head(q1_2019)
head(q1_2020)
```

# Rename columns of q1_2019 to ensure consistency
```{r}
(q1_2019 <-rename(q1_2019
                 ,ride_id=trip_id
                 ,rideable_type=bikeid
                 ,started_at=start_time
                 ,ended_at=end_time
                 ,start_station_name=from_station_name
                 ,start_station_id=from_station_id
                 ,end_station_name=to_station_name
                 ,end_station_id=to_station_id
                 ,member_casual=usertype
                 ))
```

# Observe structure of both data sets
```{r}
str(q1_2019)
str(q1_2020)
```

# Convert **ride_id** and **rideable_type** columns in ***q1_2019*** to text(character) format
```{r}
q1_2019 <- mutate(q1_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
```

# Combine both datasets into new data frame ***all_trips***
```{r}
all_trips <- bind_rows(q1_2019, q1_2020)
```

# Inspect new dataframe
```{r}
head(all_trips)
names(all_trips) <- make.names(names(all_trips))
```

# Remove unwanted columns to ensure consistency
```{r}
all_trips <- all_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng,gender,birthyear,tripduration))
```

# Inspect the column **member_casual**
```{r}
table(all_trips$member_casual)
```

# In **member_casual** column, replace *Subscriber* and *Customer* with member and casual respectively
```{r}
all_trips <-all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber"="member"
                                ,"Customer"="casual"))
```

# Add new columns
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date),"%m")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

# Add another column named **ride_length**
```{r}
all_trips <- all_trips %>% 
  mutate(ride_length= difftime(ended_at, started_at, units = "secs"))
```

# convert **ride_length** from factor to numeric to perform calculations
```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```

# Now delete some rows where bikes were taken out for quality control and ride length was negative.
# create new dataframe ***all_trips_v2***
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length <0),]
```

# Analyze ridership data by type and weekday
# create another dataframe ***customer***
```{r}
customer <- all_trips_v2 %>% 
       group_by (day_of_week, member_casual) %>%
       summarise(avg_ride_length = mean(ride_length),
                 max_ride_length = max(ride_length), 
                 min_ride_length = min(ride_length), 
                 med_ride_length = median(ride_length), 
                 num_rides = n())
```

# Rename column day_of_week
```{r}
customer<- rename(customer,
                  weekdays = day_of_week)
```

# Visualize data to compare **average ride length**
```{r}
ggplot(data = customer) +
  geom_col(aes(x=weekdays, y=avg_ride_length, fill=member_casual), position = "dodge") +
  labs(title = "Average Ride Length", subtitle = "Member vs Casual")
```

# Now visualize data to compare **number of rides**
```{r}
ggplot(data = customer) +
  geom_col(aes(x=weekdays, y=num_rides, fill=member_casual), position = "dodge") +
  labs(title = "Number of Rides", subtitle = "Member vs Casual") +
  scale_y_continuous(labels = comma)
```

# Create a dataframe **stations** for top stations with most number of rides by casual users
```{r}
stations <- all_trips_v2 %>% 
  drop_na() %>% 
  group_by(start_station_name, end_station_name, member_casual) %>% 
  summarise(total_rides = n(), .groups = 'drop') %>% 
  arrange(member_casual, -total_rides) %>% 
  slice_head(n=5)
```

# Now visualize this data by column chart
```{r}
ggplot(data=(stations)) + geom_col(aes(x=start_station_name, y=total_rides, fill = start_station_name)) +
  labs(title = "Most Popular Start Stations for Bike Rides", subtitle = "Top 3 Start Stations by Total Number of Rides Across Casual Users")

```

